Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2015-01-23 pituba/article/article_spe.tex
news+

The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system. For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production so that a pseudo-gas component can be defined. Similarly, a pseudo-oil component can be defined by subtracting the pseudo-gas component from oil phase composition. After defining the pseudo-components, the number of transport equations needed for hydrocarbon components is reduced from a usually large number to only two.

@@

 The same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}, is also applied here. This approach extends the material balance equation to saturated reservoirs under pressure maintenance (e.g. fluid injection). The applicability, however, requires negligible non-equilibrium effects in chemical equilibrium.

@@

We assume that the reservoir is a hydraulically connected volume where hydrocarbon saturation is nonzero. By a hidraulically connected volume, we mean that it is possible to observe a common pressure decline trend when any reservoir portion is under production \citep{dakepractice}. Reservoir fluids are composed of oil and gas pseudo-components and water component. There can be three fluid phases: oil, gas and (liquid) water. Oil phase is composed of oil and gas components, gas phase is composed uniquely by gas component and, water phase, uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases, i.e., the gas phase composition is always that of a dry gas.

@@

Another simplification is now assuming thermal equilibrium in the reservoir volume, i.e.,
\begin{equation}
T_w = T_o = T_g = T_{rock} = T = \text{constant} \, .
\end{equation}
Temperature is a constant prescribed value and no equation for energy balance is needed.

@@

It follows from Duhem's theorem that the equilibrium of oil and gas phases is determined if, in addition to the mass balance equations, two other properties are prescribed. From Gibbs' phase rule, the system has two degrees of freedom. It is then necessary to specify two intensive properties: the constant temperature is one and the local pressure is chosen as the other. Oil phase density now becomes a function of pressure alone. The same is valid for mass fractions of gas and oil components in oil phase:
\begin{equation}
\rho_o = \rho_o \left(p_o\right) \, , \quad Y_g = Y_g (p_o) \, , \quad Y_o = Y_o (p_o) \, .
\end{equation}

@@

Integrating Eqs.~\eqref{eq: Sw1}, \eqref{eq: So1} and \eqref{eq: Sg1} in volume and applying divergence theoreom gives:
\begin{itemize}
\item water component:
\begin{align}\label{eq: Sw2}
\begin{split}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV &= - \oint_{\partial \Omega} \rho_w \left( \vec{u}_w \cdot \vec{dS} \right) \\
&= \rho_w^{ST}\left(Q_{we}^{ST} + Q_{wi}^{ST}-Q_w^{ST}\right) \, ;
\end{split}
\end{align}

@@

We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes. In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to its volume at surface condition, i.e.:

@@

or using $\bar{Y}_o + \bar{Y}_g = 1$:
\begin{equation}
R_s = \frac{\rho_o^{ST}}{\rho_g^{ST}}\frac{\bar{Y}_g}{1-\bar{Y}_g} \, .
\end{equation}

@@

An equation for the bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure, or $R_s(p=P_b)$:
\begin{equation}\label{eq: eqPb}
\begin{split}
R_{s2}\left(p=P_b\right) = &\Big[N\left(m B_{1}/B_{g1} +R_{s1}\right)\\
&+G_i-G_p\Big]/\left(N-N_p\right) \, .
\end{split}
\end{equation}

@@

\section{Solution of material balance equation}
A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We start by defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
\begin{equation}\label{eq: residual}
\varepsilon(p_2) = \frac{E+F_i+C-F_p}{V_{p1}} \, .
\end{equation}
We proceed showing that $\varepsilon$ is a monotonic function provided some conditions on fluid properties are observed and that, as a consequence, only one root of $\varepsilon$ exists at most. Equivalently, we can say that only one solution exists for the material balance equation.

@@

The residual function clearly satisfies the first two affirmatives because phase properties are continuous everywhere and are also differentiable everywhere except for the phase change at bubble-point pressure.
The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels, each corresponding to a different phase (three fluid phases and the solid rock). For the fluid phases, each parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
\begin{equation}
\begin{split}
&V_{p1}\frac{d\varepsilon}{dp}=\underbrace{\left(N-N_p\right) \left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right)}_\text{oil phase}\\
&+\underbrace{\left[N R_{s1}+mN\frac{B_{o1}}{B_{g1}}+G_i-\left(N-N_p\right)R_{s2}-G_p\right] \frac{dB_g}{dp}}_\text{gas phase}\\
&+\underbrace{\left(W + W_i+W_e-W_p\right)\frac{dB_w}{dp}}_\text{water phase}\\
&\underbrace{-\frac{dV_p}{dp}}_\text{porous volume}\, .
\end{split}
\end{equation}

\textit{In-situ} volumes (e.g. $N-N_p$ for oil phase) are non-negative values,  pressure derivatives of fluid volumes are non-positive values (fluid volume cannot increase as pressure increases) and the pressure derivative of porous volume is a positive value (as fluid pressure increases, porous volume expands owing to rock compression).

The value of $d\varepsilon/dp$ is thus negative if at least one phase is compressible because it is a sum of products between negative and positive factors.

>>>>

The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system. For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production so that a pseudo-gas component can be defined. Similarly, a pseudo-oil component can be defined by subtraction of the pseudo-gas component from oil phase composition. After defining the pseudo-components, the number of transport equations needed for hydrocarbon components is reduced from a usually large number to only two.

@@

 The same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}, is also applied here. This approach extends the material balance equation to saturated reservoirs under pressure maintenance (e.g. fluid injection). The applicability, however, requires negligible non-equilibrium effects.

@@

We assume that the reservoir is a hydraulically connected volume where hydrocarbon saturation is nonzero. By a hidraulically connected volume, we mean that it is possible to observe a common pressure decline trend when any reservoir portion is under production \citep{dakepractice}. Reservoir fluids are composed of oil and gas pseudo-components and water component. There can be three fluid phases at most: oil, gas and (liquid) water. Oil phase is composed of oil and gas components, gas phase is composed uniquely by gas component and, water phase, uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases. The gas phase composition is always that of a dry gas.

@@

Another simplification is now made assuming thermal equilibrium in the reservoir volume, i.e.,
\begin{equation}
T_w = T_o = T_g = T_{rock} = T = \text{constant} \, .
\end{equation}
Temperature is a constant prescribed value and no equation for energy balance is needed.

@@

It follows from Duhem's theorem that the equilibrium of oil and gas phases is determined if, in addition to the mass balance equations, two other properties are prescribed. From Gibbs' phase rule, the system has two degrees of freedom. It is then necessary to specify two intensive properties: the constant temperature is one and the local pressure is chosen as the other. Since temperature was assumed constant, oil phase density becomes a function of pressure alone. The same is valid for mass fractions of gas and oil components in oil phase:
\begin{equation}
\rho_o = \rho_o \left(p_o\right) \, , \quad Y_g = Y_g (p_o) \, , \quad Y_o = Y_o (p_o) \, .
\end{equation}

@@

Integrating Eqs.~\eqref{eq: Sw1}, \eqref{eq: So1} and \eqref{eq: Sg1} in volume and applying divergence theorem gives:
\begin{itemize}
\item water component:
\begin{align}\label{eq: Sw2}
\begin{split}
\frac{\partial}{\partial t} \int\limits_{\Omega} \rho_w S_w \phi dV &= - \oint_{\partial \Omega} \rho_w \left( \vec{u}_w \cdot \vec{dS} \right) \\
&= \rho_w^{ST}\left(Q_{we}^{ST} + Q_{wi}^{ST}-Q_w^{ST}\right) \, ;
\end{split}
\end{align}

@@

We now introduce usual definitions from petroleum engineering for the description of volumetric and solubility changes. In what concerns volumetric changes, $B_w$, $B_o$ and $B_g$, the formation volume factors of water, oil and gas phases, respectively, express the ratio of a phase volume at reservoir condition to its volume at standard surface condition, i.e.:

@@

or, using $\bar{Y}_o + \bar{Y}_g = 1$:
\begin{equation}
R_s = \frac{\rho_o^{ST}}{\rho_g^{ST}}\frac{\bar{Y}_g}{1-\bar{Y}_g} \, .
\end{equation}

@@

An equation for the bubble-point can be obtained by setting $\bar{S}_{g2}=0$ in Eq.~\eqref{eq: Sg5} and solving it for the solubility ratio at bubble-point pressure, $R_s(p=P_b)$:
\begin{equation}\label{eq: eqPb}
\begin{split}
R_{s2}\left(p=P_b\right) = &\Big[N\left(m B_{1}/B_{g1} +R_{s1}\right)\\
&+G_i-G_p\Big]/\left(N-N_p\right) \, .
\end{split}
\end{equation}

@@

\section{Solution of material balance equation}
A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We start by defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
\begin{equation}\label{eq: residual}
\varepsilon(p_2) = \frac{E+F_i+C-F_p}{V_{p1}} \, .
\end{equation}
We proceed showing that $\varepsilon$ is a monotonic function provided some conditions on fluid properties are observed and that only one root of $\varepsilon$ exists at most. Equivalently, we can say that only one solution exists for the material balance equation.

@@

The residual function clearly satisfies the first two affirmatives because phase properties are continuous everywhere and are also differentiable everywhere except  at bubble-point pressure.

The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four negative terms:
\begin{equation}\label{eq: deps}
\begin{split}
&V_{p1}\frac{d\varepsilon}{dp}=\underbrace{\left(N-N_p\right) \left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right)}_\text{oil phase}\\
&+\underbrace{\left[N R_{s1}+mN\frac{B_{o1}}{B_{g1}}+G_i-\left(N-N_p\right)R_{s2}-G_p\right] \frac{dB_g}{dp}}_\text{gas phase}\\
&+\underbrace{\left(W + W_i+W_e-W_p\right)\frac{dB_w}{dp}}_\text{water phase}\\
&\underbrace{-\frac{dV_p}{dp}}_\text{porous volume}\, .
\end{split}
\end{equation}

Each term in Eq. \eqref{eq: deps} corresponds to a different phase (three fluid phases and the solid rock). For the fluid phases, each term is the product of the current volume \textit{in-situ} (a positive quantity) and a volume derivative with respect to pressure (a negative quantity). Since pore volume expands with an increase of fluid pressure, $dV_p/dp$ is positive. The value of $d\varepsilon/dp$ is thus necessarily negative.
